Taking Dyads Seriously

### 2018-02-11 paper/3_sim.tex

We utilize a simulation study to highlight the utility of AME as an inferential tool for dyadic analysis. Most scholars working with dyadic data are primarily concerned with understanding the effect of a particular independent variable on a dyadic dependent variable. The goal of the simulation is to assess how well AME can provide unbiased and well-calibrated estimates of coefficient parameters in the presence of unobserved dependencies. Specifically, we are concerned with conducting inference on regression parameters of a linear model for a network in the case where there is an omitted variable. For instance, assume that the true data-generating process for a particular $Y$ is given by:

@@

\begin{itemize}
	\item the ``standard'' international relations approach assuming independent errors; 
	\item the AME approach outlined in the previous section with a unidimensional latent factor space ($K=1$);\footnote{Results with higher values of $K$ are similar.}
	\item and an ``oracle'' regression model that assumes we have measured all sources of dependencies and thus includes both $x_{i,j}$ and $w_{i,j}$. 
\end{itemize}

>>>>

We utilize a simulation study to highlight the utility of AME as an inferential tool for dyadic analysis.\footnote{Alternative network based approaches for dyadic data are exponential random graph models (ERGMs) and the related stochastic actor oriented model (SAOM). While both these models have led to numerous contributions to a variety of literatures, the applicability of these approaches may be limited to certain types of networks and individual level characteristics. Specifically, \citet{block:etal:2017} note that these types of  models may not be appropriate in situations where network and behavioral data depend on unobserved latent variables, which is explicitly the focus of our analysis here.} Most scholars working with dyadic data are primarily concerned with understanding the effect of a particular independent variable on a dyadic dependent variable. The goal of the simulation is to assess how well AME can provide unbiased and well-calibrated estimates of coefficient parameters in the presence of unobserved dependencies. Specifically, we are concerned with conducting inference on regression parameters of a linear model for a network in the case where there is an unaccounted for higher order dependence pattern. For instance, assume that the true data-generating process for a particular $Y$ is given by:

@@

\begin{itemize}
	\item the ``standard'' international relations approach estimated through a typical generalized linear model; 
	\item the AME approach outlined in the previous section with a unidimensional latent factor space ($K=1$);\footnote{Results with higher values of $K$ are similar.}
	\item and an ``oracle'' regression model that assumes we have measured all sources of dependencies and thus includes both $x_{i,j}$ and $w_{i,j}$. 
\end{itemize}

### 2018-02-11 paper/4_app.tex

We chose to demonstrate the capabilities of AME with reference
to existing studies in order to highlight several features of the
AME approach. Firstly, the results of AME estimation are interpretable
alongside GLM results but have the additional benefit of being able to
quantify higher-order features of interest to scsholars, such as
``sender'' and ``receiver'' effects. Secondly, although the data
requirements of AME are, admittedly, more intensive than needed for GLMs, ever richer data
is increasingly prevalent in both new and existing data. Thirdly,
through AME, we can examine the extent to which interdependencies
within typical dyadic data alter previous conclusions about
international relations from recent empirical studies using a
framework that a priori assumes conditional independence.

The five studies replicated below were selected based on how recent they were and
whether they had more then 100 citations.\footnote{Note that we chose
  papers with at least 100 citations as an indicator of influence
  within the discipline. Of course, by the criteria of influence many
  other papers could have been considered for the
  replication. However, as described in the following sections, the
  AME model has specific data requirements, which limited the scope of
  potential papers.}  Each of these pieces, published in prominent
journals well-known in their respective literatures, posited a theory
in which interdependencies are consequential. However, reflecting the
dominant approach in the literature, each of the authors tested their
hypothesis by employing some form of a general linearized
regression.

In the following section, we present two sets of results for each
study, one that replicates the original study and another that tests
the central hypothesis within an AME implementation. We demonstrate
that AME allows scholars the ability to test their hypotheses of
interest, while directly accounting for both theorized and observed interdependencies in the
data. As accounting for interdependences can change the outcome
of regression models, both in substance and in estimate, we also summarize the
predictive performance of the GLM and AME implementations. This
predictive view allows us to better compare the models across
specifications. Finally, for several of the studies, we report the
AME-derived sender and receiver effects in order to highlight
additional value of the AME framework.

@@

We obtained the data for each of these studies from their replication
archives and replicated the main results of each
article.\footnote{Without exception this was straightforward to
  accomplish, thanks to the authors'  and an increasing norm in the social sciences of
  open data sharing.}  It is worth noting that the AME is applicable
only where a full set of dyads for an outcome of interest is
observable---for example it is unsuitable for studying onsets, where
dyad-years representing conflict continuations are removed from the
sample. The AME is also not applicable to analyses of case-level data,
for example, studies that examine the decision to go to war by
particular states.

@@

\begin{table}[ht]
\centering
\caption{Here we provide a brief summary of the key variable in each
  of the five replications and a note about whether or not the
  highlighted finding remains when using our network-based approach.}
	\begin{tabular}{l p{7cm} l} \toprule
		\multirow{2}{*}{Study} & \multirow{2}{*}{Central
                  Finding} &  Remains after \\ 
			& &  accounting for dependencies? \\ \toprule
		Reiter \& Stam (2003) & Personalist Regimes Attack Democracies, Not Vice Versa & {Partially Replicates} \\ \midrule
		McDonald (2004) & Lower Trade Barriers and Higher Trade Lead to Peace & {Does Not Replicate}\\ \midrule
		Rose (2004) & WTO Membership Does not Affect Trade & {Partially Replicates}\\ \midrule
		Weeks (2012) & Bosses, Juntas, and Strongmen are more Aggressive, Machines are Not & {Does Not Replicate} \\\midrule
		Gibler (2017) & Power Parity at Time of Entry to International System Inceases Conflict & {Does Not Replicate}\\ \bottomrule
	\end{tabular}
	\label{tab:modelFindingSumm}
\end{table}

An important takeaway here is that as a result of the incompatibility
between GLM models and the underlying processes that shape
international events, many scholars are forced to make knowledge claims
based on the statistical significance of a small set of covariates, or
the differences between these covariates.  These differences may
change dramatically when interdependencies are taken into account
directly, as they do in the case of this study. %MJF: Keeping this here for now, but it reads more like a conclusion
This outcome follows from AME's ability to precisely capture higher-order
interdependencies, whereas GLM approaches implicitly ignore dyadic
interdependencies. As this is a widely-known limitation of GLM
approaches, scholars often attempt to account for clustering of
observations by including additional variables and adjusting the
standard errors of the resulting estimates. At best, this method
introduces noise and imprecision into results, and at worst can
produce misleading outcomes.   Beyond just comparing parameter
estimates, we examine how well each approach can represent the data
generating process using an out-of-sample cross validation
strategy. Specifically, for each study, we randomly divide the data
into $k=30$ sets, letting $s_{ij,t}$ be the set to which pair $ij,t$
is assigned.

@@

\noindent The result of this procedure is a set of sociomatrices $\bm \hat Y$, in which each entry $\hat y_{ij,t}$ is a predicted value obtained from using a subset of the data that does not include $y_{ij,t}$.\footnote{For more details on this type of cross-validation strategy see \citet{minhas:etal:2016:arxiv}.} We summarize the performance of the various models in Table~\ref{tab:modelPerfSumm} below. For the binary models we just provide the area under the Receiver Operator Characteristic (ROC) and Precision Recall (PR) curves. Only one of the studies here had a continuous dependent variable and for this we provide the root mean squared error (RMSE) and root median squared error (RMDSE).\footnote{More details on the performance of each of these models can be found in the Appendix.} For each of the replications, we find that the AME approach substantially outperforms the original models in terms of out-of-sample predictive performance.

@@

In the original model, the variable ``Pers/Democ Directed Dyad" (which represents a Personalist $\rightarrow$ Democractic directed dyad) is positive while the variable ``Democ/Personalist Directed Dyad'' is zero and the difference between the two coefficients is clearly distinct from zero. In our re-estimation using the AME framework, we also find that Pers-Democ directed dyad has a positive effect with zero excluded from the 95\% confidence interval while Democ-Pers directed dyad is indistinguishable from zero. Using this model, however, we can no longer conclusively say that the Pers/Democratic coefficient is larger than the Democ/Personalist one. Our re-estimation using the AME approach therefore cast some doubt on Reiter \& Stam's key claim that MIDs initiated by personalist dictatorships against democracies are more likely than MIDS initiated by democracies, though many of the effects uncovered by the authors remains robust when accounting for dyadic interdependencies. Further, the effect of most of the covariates in the literature thought to predict interstate MIDs are much closer to zero when using the AME framework.

>>>>

We chose to demonstrate the capabilities of AME with reference
to existing studies in order to highlight several features of the
AME approach. Firstly, the results of AME estimation are interpretable
alongside GLM results but have the additional benefit of being able to
quantify higher-order features of interest to scholars, such as
``sender'' and ``receiver'' effects. Secondly, although the data
requirements of AME are, admittedly, more intensive than needed for GLMs, ever richer data
is increasingly prevalent in both new and existing data. Thirdly,
through AME, we can examine the extent to which interdependencies
within typical dyadic data alter previous conclusions about
international relations from recent empirical studies using a
framework that a priori assumes conditional independence.

The five studies replicated below were selected based on how recent they were and
whether they had more then 100 citations.\footnote{Note that we chose
  papers with at least 100 citations as an indicator of influence
  within the discipline. Of course, by the criteria of influence many
  other papers could have been considered for the
  replication. However, as described in the following sections, the
  AME model has specific data requirements, which limited the scope of
  potential papers.}  Each of these pieces, published in prominent
journals well-known in their respective literatures, posited a theory
in which interdependencies are consequential. Reflecting the
dominant approach in the literature, each of the authors tested their
hypothesis by employing some form of a general linearized
regression.

In the following section, we present two sets of results for each
study, one that replicates the original study and another that tests
the central hypothesis within an AME implementation. We demonstrate
that AME allows scholars the ability to test their hypotheses of
interest, while directly accounting for both theorized and observed interdependencies in the
data. As accounting for interdependences can change the outcome
of regression models, both in substance and in estimate, we also summarize the
predictive performance of the GLM and AME implementations. This
predictive view allows us to better compare the models across
specifications (and where they differ in their coefficients, see which specification does better at modeling the underlying data-generating process). Finally, for several of the studies, we report the
AME-derived sender and receiver effects in order to highlight
additional value of the AME framework.

@@

We obtained the data for each of these studies from their replication
archives and replicated the main results of each
article.\footnote{Without exception this was straightforward to
  accomplish, thanks to the authors' transparency  and an increasing norm in the social sciences of
  open data sharing.}  It is worth noting that the AME is applicable
only where a full set of dyads for an outcome of interest is
observable---for example it is unsuitable for studying onsets, where
dyad-years representing conflict continuations are removed from the
sample. The AME is also not applicable to analyses of case-level data,
for example, studies that examine the decision to go to war by
particular states.

@@

\begin{table}[ht]
\centering
\caption{Here we provide a brief summary of the key variable in each
  of the five replications and a note about whether or not the
  highlighted finding remains when using our network-based approach.}
	\begin{tabular}{l p{7cm} l} \toprule
		\multirow{2}{*}{Study} & \multirow{2}{*}{Central
                  Finding} &  Confirmed \\ 
			& &  accounting for dependencies? \\ \toprule
		Reiter \& Stam (2003) & Personalist Regimes Attack Democracies, Not Vice Versa & {Partially Confirmed} \\ \midrule
		McDonald (2004) & Lower Trade Barriers and Higher Trade Lead to Peace & {Unconfirmed}\\ \midrule
		Rose (2004) & WTO Membership Does not Affect Trade & {Partially Confirmed}\\ \midrule
		Weeks (2012) & Bosses, Juntas, and Strongmen are more Aggressive, Machines are Not & {Unconfirmed} \\\midrule
		Gibler (2017) & Power Parity at Time of Entry to International System Inceases Conflict & {Unconfirmed}\\ \bottomrule
	\end{tabular}
	\label{tab:modelFindingSumm}
\end{table}

An important takeaway here is that %as a result of the incompatibility between GLM models and the underlying processes that shape international events, 
many scholars are forced to make knowledge claims based on the statistical significance of a small set of covariates, or
the differences between these covariates.  These differences may
change dramatically when interdependencies are taken into account
directly, as they do in the case of this study. %MJF: Keeping this here for now, but it reads more like a conclusion
This outcome follows from AME's ability to precisely capture higher-order
interdependencies, whereas GLM approaches implicitly ignore dyadic
interdependencies. As this is a widely-known limitation of GLM
approaches, scholars often attempt to account for clustering of
observations by including additional variables and adjusting the
standard errors of the resulting estimates. At best, this method
introduces noise and imprecision into results, and at worst can
produce misleading outcomes.   Beyond just comparing parameter
estimates, we examine how well each approach can represent the data
generating process using an out-of-sample cross validation
strategy. Specifically, for each study, we randomly divide the data
into $k=30$ sets, letting $s_{ij,t}$ be the set to which pair $ij,t$
is assigned.

@@

\noindent The result of this procedure is a set of sociomatrices $\bm \hat Y$, in which each entry $\hat y_{ij,t}$ is a predicted value obtained from using a subset of the data that does not include $y_{ij,t}$.\footnote{For more details on this type of cross-validation strategy see \citet{minhas:etal:2016:arxiv}.} We summarize the performance of the various models in Table~\ref{tab:modelPerfSumm} below. For the binary models we  provide the area under the Receiver Operator Characteristic (ROC) and Precision Recall (PR) curves. Only one of the studies here had a continuous dependent variable and for this we provide the root mean squared error (RMSE) and root median squared error (RMDSE).\footnote{More details on the performance of each of these models can be found in the Appendix.} For each of the replications, we find that the AME approach substantially outperforms the original models in terms of out-of-sample predictive performance.

@@

In the original model, the variable ``Pers/Democ Directed Dyad" (which represents a Personalist $\rightarrow$ Democractic directed dyad) is positive while the variable ``Democ/Personalist Directed Dyad'' is zero and the difference between the two coefficients is clearly distinct from zero. In our re-estimation using the AME framework, we also find that Pers-Democ directed dyad has a positive effect with zero excluded from the 95\% confidence interval while Democ-Pers directed dyad is indistinguishable from zero. Using this model, however, we can no longer conclusively say that the Pers/Democratic coefficient is larger than the Democ/Personalist one. Our re-estimation using the AME approach therefore cast some doubt on Reiter \& Stam's key claim that MIDs initiated by personalist dictatorships against democracies are more likely than MIDS initiated by democracies, though many of the major effects  uncovered by the authors remains robust when accounting for dyadic interdependencies. Further, the effect of most of the covariates in the literature thought to predict interstate MIDs are much closer to zero when using the AME framework suggesting that some of their effects were actually the products of interdependencies.
